RVD2: An ultra-sensitive variant detection model for low-depth targeted next-generation sequencing data

### 2013-12-11 doc/ISMB2014/RVD3.tex
ISMB2014 RVD3 folder (outline and first draft).

@@

@@

@@

\subsection{Performance with read depth}\label{sec:read_depth}

We generated receiver-operating characteristic curves (ROCs) for a range of median read depth and a range of minor allele frequencies (MAFs). For these ROC curves, we used the Bayesian test without the $\chi^2$ test. Adding the $\chi^2$ improves specificity at the expense of sensitivity. Figure~\ref{fig:ROC} shows ROC curves generated by varying the threshold $\tau$ with a fixed $\alpha=0.05$. Figure~\ref{fig:ROC}A shows ROC curves for a true 0.1\% MAF for a range of median coverage depths. At the lowest depth the sensitivity and specificity is no better than random. However, we would not expect to be able to call a 1 in 1000 variant base with a coverage of only 43. The performance improves monotonically with read depth. Figures~\ref{fig:ROC}B-C show a similar relationship between coverage depth and accuracy for higher MAFs.

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=120mm]{pdf_figs/ROC_without_chi2.pdf}
\caption{ROC curve varying read depth showing detection performance.}
\label{fig:ROC}
\end{center}
\end{figure}
% from folder '2013-11-24_six_replicates_Compute_ROC_Synthetic_Qsd_0_1'

%Figure~\ref{fig:MAF} shows the posterior mean and 95\% credible intervals for $\mu_j$ for called variant positions with $\bar{n} = 5584$ and MAF = 1.0\%. All of the called positions show a clear difference between the case and control error rates. The posterior mean estimates are all shrunken towards the global error rate parameter $\mu_0 = 0.0023$ due to the hierarchical structure of the model.
%
%\begin{figure}[h]
%\begin{center}
%\includegraphics[width=120mm]{pdf_figs/Synthetic_Mubarplot_MAF1_0_Depth5584.pdf}
%\caption{Estimated minor allele fraction for called variants in 1.0\% dilution.}
%\label{fig:MAF}
%\end{center}
%\end{figure}
%% from folder '2013-11-15_synthetic_mu_barplot_optimalT', dilution rate is 0.1, downsampling rate is 0.01.

\subsection{Empirical performance compared with other algorithms}\label{sec:comparison}

We compare the empirical performance of RVD2 to other variant calling algorithms using the synthetic DNA data sets using the false discovery rate as well as sensitivity/specificity. In a research applications, the false discovery rate is a more relevant performance metrics because the aim is generally to identify interesting variants. The sensitivity/specificity metric is more relevant in clinical applications where one is more interested in correctly calling all of the positive variants and none of the negatives. GATK, Varscan2, strelka and muTect are only able to make use of one case and cone control sample, so we provide results of RVD2 with the same data ($N=1$) for a fair comparison.

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=160mm]{pdf_figs/comparison_table_ss.pdf}
\caption{Sensitivity/Specificity comparison of RVD2 with other variant calling algorithms.}
\label{tbl:comparison_ss}
\end{center}
\end{figure}
% from folder 'rvd2\results\2013-09-19_operating_characteristics'
% rvd2 data are from folder '2013-11-12_one_replicate_optT_with_chi','2013-11-12_one_replicate_T_zero_with_chi2','2013-11-15_six_replicates_optT_with_chi2','2013-11-15_six_replicates_T_zero_with_chi2'

@@

\subsection{HCC1187 primary ductal carcinoma sample}\label{sec:brca}

@@

%%%%%%%%%%%%%%%%%
% Appendix C: Data Set Statistics
%%%%%%%%%%%%%%%%%
\section{RVD2 Estimated Parameters}\label{sec:synthetic_estimate}
The RVD2 algorithm provides estimates of model parameters and latent variables given the data. We show several of these parameters in Figure~\ref{fig:depthM}.

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=120mm]{pdf_figs/depthM.pdf}
\caption{Key parameters for RVD2 model for synthetic DNA data sets.}
\label{fig:depthM}
\end{center}
\end{figure}

% from folder '22013-11-15_depthM_plot_avg10', dilution rate is 0.1.

%%%%%%%%%%%%%%%%%
% Appendix D: Parameter Settings for Variant Calling Algorithms
%%%%%%%%%%%%%%%%%
\section{Parameter settings for other variant calling algorithms}
\paragraph{\textbf{Samtools}}We used samtools (v0.1.19) function mpileup to call variants and bcftools to save the result in standard VCF files. In mpileup, we set the -d option sufficiently high at $10^6$ to avoid truncating read deapth. Option -u was enabled to make sure the output bcf files were uncompressed.

\paragraph{\textbf{GATK}}
We used GATK (v2.1-8) UnifiedGenotyper function to detect mutations on our synthetic data following the recommended workflow. Due to some format incompatibility, we applied Picard to format read group and GATK for realignment. In UnifiedGenotyper, -ploid (Number of samples in each pool $\times$ Sample Ploidy) was set at 1 because our synthetic data is haploid; -dcov was set at $10^6$ to avoid downsampling coverage within GATK.

\paragraph{\textbf{VarScan2-mpileup}}
VarScan2 (v2.3.4) mpileup2snp is a SNP calling program which takes multi-samples from samtools mpileup pipeline.We assigned parameter -C value 50 as the synthetic data was aligned using BWA and set -d at $10^6$. In mpileup2snp, --min-var-freq, the only non-default parameter, was set low enough at $10^{-5}$ because the variant frequency can be as low as $10^{-3}$.

\paragraph{\textbf{VarScan2-somatic}}
We tested VarScan2 somatic on our synthetic dataset. The parameter --normal-purity set was at 1.00, --tumor-purity at the dilution rate. The parameter --min-var-freq was set at $10^{-5}$. We combined all the positions VarScan2-somatic called regardless the somatic status (Germline/LOH/Somatic/Unknown) to compare with performance of RVD2.

\paragraph{\textbf{Strelka and muTect}}
Since configuration and Analysis for Strelka and muTect is standardized and no parameter needs to be specified, we installed these two programs and ran them on our data set separately.

Samtools mpileup takes multiple "tumor'' replicates for variant calling, so we fed six bam files from each case replicate group to mpileup. GATK, VarScan2-mpileup accept multiple ``tumor-normal'' pair replicates so we passed six pair replicates to each algorithm. Varscan2-somatic, strelka and muTect do not accept replicate data for the ``normal" or ``tumor" bam files so we used a single bam file from each replicate group with a read depth that most closely matched the overall median depth of the replicates.

% TODO: modify the following section
GATK, samtools and VarScan2-mpileup are optimized to call genotypes on pure samples. Therefore, we expect those algorithms to perform well on the 100\% dilution (pure mutant) sample and poorly on heterogeneous samples. \citet{Stead:2013fu} showed that varscan-somatic outperformed Strelka had performance on-par with muTect in detecting a 5\% MAF for read depths between 100 and 1000. We find its performance on much lower MAF variants and across a wider range of coverage depths. 
\bibliographystyle{apalike}
\bibliography{bioinfo}
\end{document}

>>>>

(RVD3) Sensitivity analysis Bayesian model
Jeffreys prior improves robustness and accuracy of Single Nucleotide Variant detection algorithm.

@@

(RVD3)
Improper Prior: Playing no prior on M is an implicit improper prior.
Jeffreys Prior:Jeffreys prior is proposed for invariance by Harold Jeffreys (1946), and defined in terms of the Fisher information. For our ploblem,
\begin{equation}\label{eqn:JefferyPrior}
\pi\left({M}_{j}\right)= I\left({M}_{j}\right)^{\frac{1}{2}}
\end{equation}

Lognormal Prior:
In log-normal distribution, the parameters denoted $\mu$ and $\sigma$, the mean and standard deviation respectively. Log-normal prior was used on $ M_{j}$.
Incorporate lognormal prior on $M_{j}$.

@@

%%%%%%%%%%%%%%%%%%
% Sampling Mj
%%%%%%%%%%%%%%%%%%
\subsection{Sampling from $p \left( M_{j} |\mu,\sigma, \theta_{ji},\mu_j\right)$}
(RVD3) For Jeffreys prior,since ${\theta }_{ji}\sim Beta\left( {\mu }_{j},{M}_{j}\right)$, the Fisher information is given by

\begin{equation}\label{equ:JefferyInference}
I\left({M}_{j}\right)={E}_{{M}_{j}}\left[ -\frac{\delta ^{2}\log p\left(\theta _{j}|\mu_{j},M_{j}\right)}{\delta M^{2}_{j}}\right]
\end{equation}

Now to calculate the equations to acquire the Fisher information, we assume there is only one replicate,

\begin{equation}\label{eqn:Betapdf}
p\left({\theta }_{j} \right)= \frac{\Gamma \left({M}_{j} \right)}{\Gamma \left({\mu }_{j} {M}_{j}\right)\Gamma \left(( 1-{\mu }_{j}){M}_{j}\right)} {{\theta}_{j}}^{{\mu}_{j}{M}_{j}-1}{\left(1-\theta\right)_{j}}^{\left(1-{\mu}_{j}\right){M}_{j}-1}
\end{equation}

\begin{equation}\label{equ:JefferyInference1}
\begin{split}
\log p\left(\theta_{j}|\mu_{j},M_{j}\right)& =\log \Gamma \left(M_{j}\right)-\log \Gamma\left(\mu_{j},M_{j}\right)- \log \Gamma\left(1-\mu_{j},M_{j}\right)\\
& + (\mu_{j}M_{j}-1)\log\theta_{j} + ((1-\,u_{j})M_{j}-1)\log(1-\theta_{j})\
\end{split}
\end{equation}

\begin{equation}
\frac{\delta\log p(\theta_{j})}{\delta M_{j}} = \Psi(M_{j}) - \Psi(\mu_{j} M_{j})\mu_{j} - \Psi((1-\mu_{j})M_{j})(1-\mu_{j}) +\mu_{j}\log\theta_{j} + (1-\mu_{j})\log(1-\theta_{j})
\end{equation}

\begin{equation}
\frac{\delta^{2}\log p(\theta_{j})}{\delta M_{j}^{2}}  = \Psi_{1}(M_{j}) - \Psi_{1}(\mu_{j} M_{j})\mu_{j}^{2} - \Psi_{1}((1-\mu_{j})M_{j})(1-\mu_{j})^{2}
\end{equation}

Now we have the Jeffreys' prior for $M_{j}$:

\begin{equation}
\pi\left({M}_{j}\right) = [-\left(\Psi_{1}(M_{j}) - \Psi_{1}(\mu_{j} M_{j})\mu_{j}^{2} - \Psi_{1}((1-\mu_{j})M_{j}){(1-\mu_{j})^{2}}\right)]^{\frac{1}{2}}
\end{equation}

For log-normal prior, the posterior distribution over $M_{j}$ given its Markov blanket is

\begin{equation}
	p( M_{j} |\mu, \sigma, \theta_{ji},\mu_j) \propto p(\theta_{ji} | \mu_j, M_j) p(M_{j} | \mu, \sigma)
\end{equation}

We have $ \theta_{ji}\thicksim\text{Beta}(\mu_{j},M_{j})$, and $ M_{j} \thicksim \text{log-normal}(\mu, \sigma)$. Since we cannot give an analytical form for this, so we sample from the posterior distribution using the Metropolis-Hastings algorithm.

@@

\subsection{Sensitivity analysis of priors}
In order to examine the robustness of RVD3 to changes of different priors, we...

We found that RVD3 is robust to changing the priors.
The RVD3 algorithm provides estimates of model parameters and latent variables given the data. We show several of these parameters in Figure~\ref{fig:M_loc_rvd29} and Figure ~\ref{fig:M_loc_rvd30}.

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=120mm]{figs/M_loc_rvd29.pdf}
\caption{Key parameters for RVD3 model with Jeffreys prior for synthetic DNA data sets.}
\label{fig:M_loc_rvd29}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=120mm]{figs/M_loc_rvd29.pdf}
\caption{Key parameters for RVD3 model with log-normal prior for synthetic DNA data sets.}
\label{fig:M_loc_rvd30}
\end{center}
\end{figure}
% from folder '22013-11-15_depthM_plot_avg10', dilution rate is 0.1.

\subsection{Results of priors on synthetic data}
Performance with read depth:
We generated receiver-operating characteristic curves (ROCs) for a range of median read depth and a range of minor allele frequencies (MAFs). For these ROC curves, we used the Bayesian test without the $\chi^2$ test. Adding the $\chi^2$ improves specificity at the expense of sensitivity. Figure~\ref{fig:ROC} shows ROC curves generated by varying the threshold $\tau$ with a fixed $\alpha=0.05$. Figure~\ref{fig:ROC}A shows ROC curves for a true 0.1\% MAF for a range of median coverage depths. At the lowest depth the sensitivity and specificity is no better than random. However, we would not expect to be able to call a 1 in 1000 variant base with a coverage of only 43. The performance improves monotonically with read depth. Figures~\ref{fig:ROC_29}B-C show a similar relationship between coverage depth and accuracy for higher MAFs.

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=120mm]{figs/ROC_without_chi2_rvd29.pdf}
\caption{ROC curve varying read depth showing detection performance.}
\label{fig:ROC_29}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=120mm]{figs/ROC_without_chi2_rvd30.pdf}
\caption{ROC curve varying read depth showing detection performance.}
\label{fig:ROC_30}
\end{center}
\end{figure}

Empirical performance compared with other algorithms:
We compare the empirical performance of RVD2 to other variant calling algorithms using the synthetic DNA data sets using the false discovery rate as well as sensitivity/specificity. In a research applications, the false discovery rate is a more relevant performance metrics because the aim is generally to identify interesting variants. The sensitivity/specificity metric is more relevant in clinical applications where one is more interested in correctly calling all of the positive variants and none of the negatives. GATK, Varscan2, strelka and muTect are only able to make use of one case and cone control sample, so we provide results of RVD2 with the same data ($N=1$) for a fair comparison.

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=160mm]{tables/comparison_table_ss.pdf}
\caption{Sensitivity/Specificity comparison of RVD3 with different priors.}
\label{tbl:table_29}
\end{center}
\end{figure}

@@

\subsection{Results of Jeffreys prior on yeast data}

@@

\bibliographystyle{apalike}
\bibliography{bioinfo}
\end{document}
